---
title: "Final Project"
author: "Prabal Ghosh"
date: "January 31, 2019"
output: html_document
---

##Background

##Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

##In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).
##Dataset Overview
##A short description of the datasets content from the authors' website:

##Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

##Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg)
##Environment Preparation
##We first upload the R libraries that are necessary for the complete analysis
```{r}
library(knitr)
```
```{r}
library(caret)
```

```{r}
library(rpart)
```
```{r}
library(rpart.plot)
```
```{r}
library(rattle)
```

```{r}
library(randomForest)
```

```{r}
set.seed(12345)
```
##Goal

##The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases

##Loading the data

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```
```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
```

```{r}
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```
##Partioning the training set into two

```{r}
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
```

```{r}
myTraining <- training[inTrain, ]
```

```{r}
myTesting <- training[-inTrain, ]
```

```{r}
dim(myTraining); dim(myTesting)
```

## Cleaning the data

##Remove NearZeroVariance variables

```{r}
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]
```

##Remove the first column of the myTraining data set

```{r}
myTraining <- myTraining[c(-1)]
```

##Clean variables with more than 60% NA

```{r}
trainingPra <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingPra)) {
            if( length( grep(names(myTraining[i]), names(trainingPra)[j]) ) == 1)  {
                trainingPra <- trainingPra[ , -j]
            }   
        } 
    }
}

```
```{r}
myTraining <- trainingPra
```

```{r}
rm(trainingPra)
```
##Transform the myTesting and testing data sets

```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])
myTesting <- myTesting[clean1]
testing <- testing[clean2]
```

```{r}
dim(myTesting)
```

##Change the data into the same type
```{r}
for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
```

```{r}
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]
```
##Prediction with Decision Trees

```{r}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)
```
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree
```

```{r}
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```






##Prediction with Random Forests

```{r}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=myTraining)
predictionB1 <- predict(modFitB1, myTesting, type = "class")
cmrf <- confusionMatrix(predictionB1, myTesting$classe)
cmrf
```

```{r}
plot(modFitB1)
```
```{r}
plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

##Final Result on the test data

##Random Forests gave an Accuracy in the myTesting dataset of 99.86%, which was more accurate that what I got from the Decision Trees
##The expected out-of-sample error is 100-99.86 = 0.14%.

```{r}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```



